---
title: "data_manipulation"
output: github_document
date: "2025-09-18"
---

```{r}
library(tidyverse)
```

Let's import the datasets

```{r}
litters_df =
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", ""))
litters_df = 
    janitor::clean_names(litters_df)

pups_df = 
    read_csv("data_import_examples/FAS_pups.csv", na = c("NA", "."), skip = 3)
pups_df = 
    janitor::clean_names(pups_df)
```

##'Select'

'Select' selects variables from my dataframe

generates new dataset with just those variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range
```{r}
select(litters_df, group:gd_of_birth)
```
colon above says keep this variable through this variable

select by removal
```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix
keep group and any var that starts with gd
```{r}
select(litters_df, group, starts_with("gd"))
```

rerrange by selection
if pair with everything fucntion, keeps litter number and everything else
relocate does same as everything function
```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())

relocate(litters_df, litter_number)
```

rename by selecting
```{r}
select(litters_df, GROUP=group, everything())
rename(litters_df, GROUP=group)
```

what if I only select one variable...?
select keeps it in its respective column

almost never use it, just pulls obs from var
```{r}
select(litters_df, group)

pull(litters_df, group)
```

learning assessment
```{r}
select(pups_df, litter_number, sex, pd_ears)
```








