---
title: "data_manipulation"
output: github_document
date: "2025-09-18"
---

```{r}
library(tidyverse)
```

Let's import the datasets

```{r}
litters_df =
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", ""))
litters_df = 
    janitor::clean_names(litters_df)

pups_df = 
    read_csv("data_import_examples/FAS_pups.csv", na = c("NA", "."), skip = 3)
pups_df = 
    janitor::clean_names(pups_df)
```

##'Select'

'Select' selects variables from my dataframe

generates new dataset with just those variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range
```{r}
select(litters_df, group:gd_of_birth)
```
colon above says keep this variable through this variable

select by removal
```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix
keep group and any var that starts with gd
```{r}
select(litters_df, group, starts_with("gd"))
```

rerrange by selection
if pair with everything fucntion, keeps litter number and everything else
relocate does same as everything function
```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())

relocate(litters_df, litter_number)
```

rename by selecting
```{r}
select(litters_df, GROUP=group, everything())
rename(litters_df, GROUP=group)
```

what if I only select one variable...?
select keeps it in its respective column

almost never use it, just pulls obs from var
```{r}
select(litters_df, group)

pull(litters_df, group)
```

learning assessment
```{r}
select(pups_df, litter_number, sex, pd_ears)
```

##fliter

use 'filter()' to filter out rows
filter gets rid of everything false
```{r}
filter(litters_df, gd_of_birth==20, pups_born_alive==6)
```

when want to look for more than five pups born alive
can also do greater than equal to or less than equal to
```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive >= 5)
filter(litters_df, pups_born_alive <= 5)
```

looking for non-matches
```{r}
filter(litters_df, pups_born_alive != 6)
```

match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing rows with dropna(), instead of using fikter

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight) #only dropping missingness for columns don't want missing data in
```

```{r}
filter(pups_df, sex==1)
filter(pups_df, pd_walk < 11, sex==2)
```

##'mutate'

use mutate() to create or modify variables.

```{r}
mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

## 'arrange'

arrange() arranges things

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, group, pups_born_alive)

arrange(litters_df, gd0_weight)
```


## PIPES

Pipe operator starts to pull stuff together

Piping solves this problem. It allows you to turn the nested approach into a sequential chain by passing the result of one function call as an argument to the next function call:

"command shift m" gives pipe operator
|> means AND THEN...
assumes that whatever comes out of pipe then goes into the next automatically; already knows which dataframe to use

below, reads csv and then cleans names and then selects
```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", "")) |>  #and then clean
  janitor::clean_names() |> #and then select variables
  select(group, litter_number, starts_with("gd")) |> #and then drop missing values
  drop_na() |> #and then create new var weight gain
  mutate(
    wt_gain = gd18_weight - gd0_weight,
   )
view(litters_df)
```

learning assessment:
import and tidy the pups data in a piped chain of commands
```{r}
pups_df =
  read_csv("data_import_examples/FAS_pups.csv", na = c("NA", ".", ""), skip=3) |>
  janitor::clean_names() |> 
  filter(sex==1) |> 
  select(-pd_ears) |> 
  mutate(
    Pd_pivot_ge7 = pd_pivot >= 7
  )

view(pups_df)
  
```




















